---
title: "LengthCompAnalysis"
author: 'Mohith Reddy Arikatla (CWID: moa4020)'
date: "2024-02-09"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(fitdistrplus)
library(logspline)
library(dplyr)
library(DESeq2)
library(magrittr)
library(pheatmap)
library(ggplot2)
library(tidyverse)
library(mnormt)
library(nlcor)
library(RColorBrewer)
```

```{r}
gencode.vM33.All.GeneLengths <- read.table(file="gencode.vM33.All.GeneLengths.txt",sep="\t",col.names = c("GeneID","GeneLength"))

# Convert to Numeric
gencode.vM33.All.GeneLengths <- gencode.vM33.All.GeneLengths %>%
  mutate(across(-1, as.numeric))

row.names(gencode.vM33.All.GeneLengths) <- sub("\\..*", "", gencode.vM33.All.GeneLengths$GeneID)
```

```{r}
hist(gencode.vM33.All.GeneLengths$GeneLength)

gencode.vM33.All.GeneLengths$log2.GeneLength <- log2(gencode.vM33.All.GeneLengths$GeneLength) + 1

par(mar = c(5, 5, 2, 2))
hist(gencode.vM33.All.GeneLengths$log2.GeneLength, freq = FALSE)
lines(density(gencode.vM33.All.GeneLengths$log2.GeneLength))
```

```{r }
par(mar = c(5, 5, 2, 2))
descdist(gencode.vM33.All.GeneLengths$log2.GeneLength, discrete = FALSE)
```

```{r }
fit.norm <- fitdist(gencode.vM33.All.GeneLengths$log2.GeneLength, "norm")
par(mar = c(5, 5, 2, 2))
plot(fit.norm)
```

```{r }}
fit.weibull <- fitdist(gencode.vM33.All.GeneLengths$log2.GeneLength, "weibull")
par(mar = c(5, 5, 2, 2))
plot(fit.weibull)
```

```{r }
fit.lognorm <- fitdist(gencode.vM33.All.GeneLengths$log2.GeneLength, "lnorm")
par(mar = c(5, 5, 2, 2))
plot(fit.lognorm)
```

```{r }
fit.norm$aic
fit.weibull$aic
fit.lognorm$aic
```

Choose top and bottom 5% genes and median genes in our DESeq datasets for sampling. 

```{r message=FALSE}
PDDo <- "/Users/mar/Library/CloudStorage/GoogleDrive-mohitharikatla.education@gmail.com/My Drive/RR9_MurineSequencingData/Thesis/02_rr9/00.1_Overlap_DESeq/PostDESeq_overlap.RData"
load(PDDo)
```

### Gene Length Comparison (Overlap)

```{r}
gencode.vM33.All.GeneLengths.filtered <- gencode.vM33.All.GeneLengths[Overlap_names,]
```

```{r}
median_value <- median(gencode.vM33.All.GeneLengths.filtered$log2.GeneLength)

# Calculate the absolute differences between each data point and the median
gencode.vM33.All.GeneLengths.filtered$ab_diff <- abs(gencode.vM33.All.GeneLengths.filtered$log2.GeneLength - median_value)

median_500_genes <- gencode.vM33.All.GeneLengths.filtered[order(gencode.vM33.All.GeneLengths.filtered$ab_diff)[1:500],]

top_500_genes <- gencode.vM33.All.GeneLengths.filtered[order(-gencode.vM33.All.GeneLengths.filtered$GeneLength)[1:500],]

bottom_500_genes <- gencode.vM33.All.GeneLengths.filtered[order(gencode.vM33.All.GeneLengths.filtered$GeneLength)[1:500],]
```

```{r}
sample_genes <- list(
  median500 = median_500_genes,
  top500 = top_500_genes,
  bottom500 = bottom_500_genes
)

ONT_median.ds <- ONT_DESeq_overlap.ds[row.names(median_500_genes), ]
Illumina_median.ds <- Illumina_DESeq_overlap.ds[row.names(median_500_genes), ]

ONT_top.ds <- ONT_DESeq_overlap.ds[row.names(top_500_genes), ]
Illumina_top.ds <- Illumina_DESeq_overlap.ds[row.names(top_500_genes), ]

ONT_bottom.ds <- ONT_DESeq_overlap.ds[row.names(bottom_500_genes), ]
Illumina_bottom.ds <- Illumina_DESeq_overlap.ds[row.names(bottom_500_genes), ]
```

```{r}
ONT_median.rlog <- rlog(ONT_median.ds,blind=TRUE)
Illumina_median.rlog <- rlog(Illumina_median.ds,blind=TRUE)

plotPCA(ONT_median.rlog, intgroup="condition", ntop = 100, returnData=FALSE)
plotPCA(Illumina_median.rlog, intgroup="condition", ntop = 100, returnData=FALSE)
```

```{r}
ONT_top.rlog <- rlog(ONT_top.ds,blind=TRUE)
Illumina_top.rlog <- rlog(Illumina_top.ds,blind=TRUE)

plotPCA(ONT_top.rlog, intgroup="condition", ntop = 100, returnData=FALSE)
plotPCA(Illumina_top.rlog, intgroup="condition", ntop = 100, returnData=FALSE)
```

```{r}
ONT_bottom.rlog <- rlog(ONT_bottom.ds,blind=TRUE)
Illumina_bottom.rlog <- rlog(Illumina_bottom.ds,blind=TRUE)

plotPCA(ONT_bottom.rlog, intgroup="condition", ntop = 100, returnData=FALSE)
plotPCA(Illumina_bottom.rlog, intgroup="condition", ntop = 100, returnData=FALSE)
```
```{r}
design(ONT_bottom.ds) <- ~ condition
ONT_bottom.ds$condition %<>% relevel(ref="GC")
ONT_bottom.ds %<>% DESeq()
# normalize for diffs in sequencing depth and abundance per sample
ONT_bottom.ds %<>% estimateSizeFactors()
# gene-wise dispersion estimates across all samples
ONT_bottom.ds %<>% estimateDispersions()
# fit a neg. binomial GLM and compute Wald stat for each gene 
ONT_bottom.ds %<>% nbinomWaldTest()

ONT_bottom.results <- results(ONT_bottom.ds,independentFiltering = TRUE,alpha = 0.05)

design(Illumina_bottom.ds) <- ~ condition
Illumina_bottom.ds$condition %<>% relevel(ref="GC")
Illumina_bottom.ds %<>% DESeq()
# normalize for diffs in sequencing depth and abundance per sample
Illumina_bottom.ds %<>% estimateSizeFactors()
# gene-wise dispersion estimates across all samples
Illumina_bottom.ds %<>% estimateDispersions()
# fit a neg. binomial GLM and compute Wald stat for each gene 
Illumina_bottom.ds %<>% nbinomWaldTest()

Illumina_bottom.results <- results(Illumina_bottom.ds,independentFiltering = TRUE,alpha = 0.05)
```

```{r}
design(ONT_top.ds) <- ~ condition
ONT_top.ds$condition %<>% relevel(ref="GC")
ONT_top.ds %<>% DESeq()
# normalize for diffs in sequencing depth and abundance per sample
ONT_top.ds %<>% estimateSizeFactors()
# gene-wise dispersion estimates across all samples
ONT_top.ds %<>% estimateDispersions()
# fit a neg. binomial GLM and compute Wald stat for each gene 
ONT_top.ds %<>% nbinomWaldTest()

ONT_top.results <- results(ONT_top.ds,independentFiltering = TRUE,alpha = 0.05)

design(Illumina_top.ds) <- ~ condition
Illumina_top.ds$condition %<>% relevel(ref="GC")
Illumina_top.ds %<>% DESeq()
# normalize for diffs in sequencing depth and abundance per sample
Illumina_top.ds %<>% estimateSizeFactors()
# gene-wise dispersion estimates across all samples
Illumina_top.ds %<>% estimateDispersions()
# fit a neg. binomial GLM and compute Wald stat for each gene 
Illumina_top.ds %<>% nbinomWaldTest()

Illumina_top.results <- results(Illumina_top.ds,independentFiltering = TRUE,alpha = 0.05)
```
```{r}
design(ONT_median.ds) <- ~ condition
ONT_median.ds$condition %<>% relevel(ref="GC")
ONT_median.ds %<>% DESeq()
# normalize for diffs in sequencing depth and abundance per sample
ONT_median.ds %<>% estimateSizeFactors()
# gene-wise dispersion estimates across all samples
ONT_median.ds %<>% estimateDispersions()
# fit a neg. binomial GLM and compute Wald stat for each gene 
ONT_median.ds %<>% nbinomWaldTest()

ONT_median.results <- results(ONT_median.ds,independentFiltering = TRUE,alpha = 0.05)

design(Illumina_median.ds) <- ~ condition
Illumina_median.ds$condition %<>% relevel(ref="GC")
Illumina_median.ds %<>% DESeq()
# normalize for diffs in sequencing depth and abundance per sample
Illumina_median.ds %<>% estimateSizeFactors()
# gene-wise dispersion estimates across all samples
Illumina_median.ds %<>% estimateDispersions()
# fit a neg. binomial GLM and compute Wald stat for each gene 
Illumina_median.ds %<>% nbinomWaldTest()

Illumina_median.results <- results(Illumina_median.ds,independentFiltering = TRUE,alpha = 0.05)
```

```{r}
ONT_median.results <- ONT_median.results %>% `[`(order(.$padj),)
head(ONT_median.results)

par(mfrow=c(1,2))
plotCounts(ONT_median.ds, gene="ENSMUSG00000029343", normalized = TRUE, xlab="")
plotCounts(ONT_median.ds, gene = which.max(ONT_median.results$padj), xlab="",
           main = "Gene with max. p.adj.\n(=least significant)")
```


```{r}
Illumina_median.results <- Illumina_median.results %>% `[`(order(.$padj),)
head(Illumina_median.results)

par(mfrow=c(1,2))
plotCounts(Illumina_median.ds, gene="ENSMUSG00000018752", normalized = TRUE, xlab="")
plotCounts(Illumina_median.ds, gene = which.max(Illumina_median.results$padj), xlab="",
           main = "Gene with max. p.adj.\n(=least significant)")
```

```{r}
# identify genes with the desired adjusted p-value cut-off
Illumina_median.DGEgenes <- rownames(subset(Illumina_median.results, padj < 0.05)) 
# extract rlog-transformed values into a matrix
Illumina_median_rlog.dge <- Illumina_median.rlog[Illumina_median.DGEgenes,] %>% assay
# heatmap of DEG sorted by p.adjust
pheatmap(Illumina_median_rlog.dge, scale="row",
         show_rownames=FALSE, main="Illumina DGE",
         color=colorRampPalette(RColorBrewer::brewer.pal(n=7, name="Reds"))(100))

# identify genes with the desired adjusted p-value cut-off
Illumina_top.DGEgenes <- rownames(subset(Illumina_top.results, padj < 0.05)) 
# extract rlog-transformed values into a matrix
Illumina_top_rlog.dge <- Illumina_top.rlog[Illumina_top.DGEgenes,] %>% assay
# heatmap of DEG sorted by p.adjust
pheatmap(Illumina_top_rlog.dge, scale="row",
         show_rownames=FALSE, main="Illumina DGE",
         color=colorRampPalette(RColorBrewer::brewer.pal(n=7, name="Reds"))(100))

# identify genes with the desired adjusted p-value cut-off
Illumina_bottom.DGEgenes <- rownames(subset(Illumina_bottom.results, padj < 0.05)) 
# extract rlog-transformed values into a matrix
Illumina_bottom_rlog.dge <- Illumina_bottom.rlog[Illumina_bottom.DGEgenes,] %>% assay
# heatmap of DEG sorted by p.adjust
# pheatmap(Illumina_bottom_rlog.dge, scale="row",
#          show_rownames=FALSE, main="Illumina DGE",
#          color=colorRampPalette(RColorBrewer::brewer.pal(n=7, name="Reds"))(100))
```
changing the color scheme on pheatmap to see more contrast
```{r}


# identify genes with the desired adjusted p-value cut-off
ONT_median.DGEgenes <- rownames(subset(ONT_median.results, pvalue<0.05)) 
# extract rlog-transformed values into a matrix
ONT_median_rlog.dge <- ONT_median.rlog[ONT_median.DGEgenes,] %>% assay
# heatmap of DEG sorted by p.adjust
pheatmap(ONT_median_rlog.dge, scale="row",
         show_rownames=FALSE, main="ONT DGE",
         color = colorRampPalette(c("blue", "red"))(100))

# identify genes with the desired adjusted p-value cut-off
ONT_top.DGEgenes <- rownames(subset(ONT_top.results, pvalue<0.05)) 
# extract rlog-transformed values into a matrix
ONT_top_rlog.dge <- ONT_top.rlog[ONT_top.DGEgenes,] %>% assay
# heatmap of DEG sorted by p.adjust
pheatmap(ONT_top_rlog.dge, scale="row",
         show_rownames=FALSE, main="ONT DGE",
         color = colorRampPalette(c("blue", "red"))(100))

# identify genes with the desired adjusted p-value cut-off
ONT_bottom.DGEgenes <- rownames(subset(ONT_bottom.results, pvalue<0.05)) 
# extract rlog-transformed values into a matrix
ONT_bottom_rlog.dge <- ONT_bottom.rlog[ONT_bottom.DGEgenes,] %>% assay
# heatmap of DEG sorted by p.adjust
pheatmap(ONT_bottom_rlog.dge, scale="row",
         show_rownames=FALSE, main="ONT DGE",
         color = colorRampPalette(c("blue", "red"))(100))
```
Since nanopore is not giving a great result with 500 genes. Let's try increasing the gene set size. 

```{r}

median_1000_genes <- gencode.vM33.All.GeneLengths.filtered[order(gencode.vM33.All.GeneLengths.filtered$ab_diff)[1:1000],]

top_1000_genes <- gencode.vM33.All.GeneLengths.filtered[order(-gencode.vM33.All.GeneLengths.filtered$GeneLength)[1:1000],]

bottom_1000_genes <- gencode.vM33.All.GeneLengths.filtered[order(gencode.vM33.All.GeneLengths.filtered$GeneLength)[1:1000],]

sample_genes <- list(
  median500 = median_500_genes,
  top500 = top_500_genes,
  bottom500 = bottom_500_genes,
  median1000 = median_1000_genes,
  top1000 = top_1000_genes,
  bottom1000 = bottom_1000_genes
)

ONT_median1000.ds <- ONT_DESeq_overlap.ds[row.names(median_1000_genes), ]
ONT_top1000.ds <- ONT_DESeq_overlap.ds[row.names(top_1000_genes), ]
ONT_bottom1000.ds <- ONT_DESeq_overlap.ds[row.names(bottom_1000_genes), ]

ONT_median1000.rlog <- rlog(ONT_median1000.ds,blind=TRUE)
plotPCA(ONT_median1000.rlog, intgroup="condition", ntop = 100, returnData=FALSE)

ONT_top1000.rlog <- rlog(ONT_top1000.ds,blind=TRUE)
plotPCA(ONT_top1000.rlog, intgroup="condition", ntop = 100, returnData=FALSE)

ONT_bottom1000.rlog <- rlog(ONT_bottom1000.ds,blind=TRUE)
plotPCA(ONT_bottom1000.rlog, intgroup="condition", ntop = 100, returnData=FALSE)

design(ONT_median1000.ds) <- ~ condition
ONT_median1000.ds$condition %<>% relevel(ref="GC")
ONT_median1000.ds %<>% DESeq()
# normalize for diffs in sequencing depth and abundance per sample
ONT_median1000.ds %<>% estimateSizeFactors()
# gene-wise dispersion estimates across all samples
ONT_median1000.ds %<>% estimateDispersions()
# fit a neg. binomial GLM and compute Wald stat for each gene 
ONT_median1000.ds %<>% nbinomWaldTest()

ONT_median1000.results <- results(ONT_median1000.ds,independentFiltering = TRUE,alpha = 0.05)

design(ONT_top1000.ds) <- ~ condition
ONT_top1000.ds$condition %<>% relevel(ref="GC")
ONT_top1000.ds %<>% DESeq()
# normalize for diffs in sequencing depth and abundance per sample
ONT_top1000.ds %<>% estimateSizeFactors()
# gene-wise dispersion estimates across all samples
ONT_top1000.ds %<>% estimateDispersions()
# fit a neg. binomial GLM and compute Wald stat for each gene 
ONT_top1000.ds %<>% nbinomWaldTest()

ONT_top1000.results <- results(ONT_top1000.ds,independentFiltering = TRUE,alpha = 0.05)


design(ONT_bottom1000.ds) <- ~ condition
ONT_bottom1000.ds$condition %<>% relevel(ref="GC")
ONT_bottom1000.ds %<>% DESeq()
# normalize for diffs in sequencing depth and abundance per sample
ONT_bottom1000.ds %<>% estimateSizeFactors()
# gene-wise dispersion estimates across all samples
ONT_bottom1000.ds %<>% estimateDispersions()
# fit a neg. binomial GLM and compute Wald stat for each gene 
ONT_bottom1000.ds %<>% nbinomWaldTest()

ONT_bottom1000.results <- results(ONT_bottom1000.ds,independentFiltering = TRUE,alpha = 0.05)

# identify genes with the desired adjusted p-value cut-off
ONT_median1000.DGEgenes <- rownames(subset(ONT_median1000.results, pvalue<0.05)) 
# extract rlog-transformed values into a matrix
ONT_median1000_rlog.dge <- ONT_median1000.rlog[ONT_median1000.DGEgenes,] %>% assay
# heatmap of DEG sorted by p.adjust
pheatmap(ONT_median1000_rlog.dge, scale="row",
         show_rownames=FALSE, main="ONT DGE",
         color = colorRampPalette(c("blue", "red"))(100))

# identify genes with the desired adjusted p-value cut-off
ONT_top1000.DGEgenes <- rownames(subset(ONT_top1000.results, pvalue<0.05)) 
# extract rlog-transformed values into a matrix
ONT_top1000_rlog.dge <- ONT_top1000.rlog[ONT_top1000.DGEgenes,] %>% assay
# heatmap of DEG sorted by p.adjust
pheatmap(ONT_top1000_rlog.dge, scale="row",
         show_rownames=FALSE, main="ONT DGE",
         color = colorRampPalette(c("blue", "red"))(100))

# identify genes with the desired adjusted p-value cut-off
ONT_bottom1000.DGEgenes <- rownames(subset(ONT_bottom1000.results, pvalue<0.05)) 
# extract rlog-transformed values into a matrix
ONT_bottom1000_rlog.dge <- ONT_bottom1000.rlog[ONT_bottom1000.DGEgenes,] %>% assay
# heatmap of DEG sorted by p.adjust
pheatmap(ONT_bottom1000_rlog.dge, scale="row",
         show_rownames=FALSE, main="ONT DGE",
         color = colorRampPalette(c("blue", "red"))(100))
```

```{r}

median_5000_genes <- gencode.vM33.All.GeneLengths.filtered[order(gencode.vM33.All.GeneLengths.filtered$ab_diff)[1:5000],]

top_5000_genes <- gencode.vM33.All.GeneLengths.filtered[order(-gencode.vM33.All.GeneLengths.filtered$GeneLength)[1:5000],]

bottom_5000_genes <- gencode.vM33.All.GeneLengths.filtered[order(gencode.vM33.All.GeneLengths.filtered$GeneLength)[1:5000],]

sample_genes <- list(
  median500 = median_500_genes,
  top500 = top_500_genes,
  bottom500 = bottom_500_genes,
  median1000 = median_1000_genes,
  top1000 = top_1000_genes,
  bottom1000 = bottom_1000_genes,
  median5000 = median_5000_genes,
  top5000 = top_5000_genes,
  bottom5000 = bottom_5000_genes
)

ONT_median5000.ds <- ONT_DESeq_overlap.ds[row.names(median_5000_genes), ]
ONT_top5000.ds <- ONT_DESeq_overlap.ds[row.names(top_5000_genes), ]
ONT_bottom5000.ds <- ONT_DESeq_overlap.ds[row.names(bottom_5000_genes), ]

ONT_median5000.rlog <- rlog(ONT_median5000.ds,blind=TRUE)
plotPCA(ONT_median5000.rlog, intgroup="condition", ntop = 100, returnData=FALSE)

ONT_top5000.rlog <- rlog(ONT_top5000.ds,blind=TRUE)
plotPCA(ONT_top5000.rlog, intgroup="condition", ntop = 100, returnData=FALSE)

ONT_bottom5000.rlog <- rlog(ONT_bottom5000.ds,blind=TRUE)
plotPCA(ONT_bottom5000.rlog, intgroup="condition", ntop = 100, returnData=FALSE)

design(ONT_median5000.ds) <- ~ condition
ONT_median5000.ds$condition %<>% relevel(ref="GC")
ONT_median5000.ds %<>% DESeq()
# normalize for diffs in sequencing depth and abundance per sample
ONT_median5000.ds %<>% estimateSizeFactors()
# gene-wise dispersion estimates across all samples
ONT_median5000.ds %<>% estimateDispersions()
# fit a neg. binomial GLM and compute Wald stat for each gene 
ONT_median5000.ds %<>% nbinomWaldTest()

ONT_median5000.results <- results(ONT_median5000.ds,independentFiltering = TRUE,alpha = 0.05)

design(ONT_top5000.ds) <- ~ condition
ONT_top5000.ds$condition %<>% relevel(ref="GC")
ONT_top5000.ds %<>% DESeq()
# normalize for diffs in sequencing depth and abundance per sample
ONT_top5000.ds %<>% estimateSizeFactors()
# gene-wise dispersion estimates across all samples
ONT_top5000.ds %<>% estimateDispersions()
# fit a neg. binomial GLM and compute Wald stat for each gene 
ONT_top5000.ds %<>% nbinomWaldTest()

ONT_top5000.results <- results(ONT_top5000.ds,independentFiltering = TRUE,alpha = 0.05)


design(ONT_bottom5000.ds) <- ~ condition
ONT_bottom5000.ds$condition %<>% relevel(ref="GC")
ONT_bottom5000.ds %<>% DESeq()
# normalize for diffs in sequencing depth and abundance per sample
ONT_bottom5000.ds %<>% estimateSizeFactors()
# gene-wise dispersion estimates across all samples
ONT_bottom5000.ds %<>% estimateDispersions()
# fit a neg. binomial GLM and compute Wald stat for each gene 
ONT_bottom5000.ds %<>% nbinomWaldTest()

ONT_bottom5000.results <- results(ONT_bottom5000.ds,independentFiltering = TRUE,alpha = 0.05)

# identify genes with the desired adjusted p-value cut-off
ONT_median5000.DGEgenes <- rownames(subset(ONT_median5000.results, padj<0.05)) 
# extract rlog-transformed values into a matrix
ONT_median5000_rlog.dge <- ONT_median5000.rlog[ONT_median5000.DGEgenes,] %>% assay
# heatmap of DEG sorted by p.adjust
pheatmap(ONT_median5000_rlog.dge, scale="row",
         show_rownames=FALSE, main="ONT DGE",
         color = colorRampPalette(c("blue", "red"))(100))

# identify genes with the desired adjusted p-value cut-off
ONT_top5000.DGEgenes <- rownames(subset(ONT_top5000.results, padj<0.05)) 
# extract rlog-transformed values into a matrix
ONT_top5000_rlog.dge <- ONT_top5000.rlog[ONT_top5000.DGEgenes,] %>% assay
# heatmap of DEG sorted by p.adjust
pheatmap(ONT_top5000_rlog.dge, scale="row",
         show_rownames=FALSE, main="ONT DGE",
         color = colorRampPalette(c("blue", "red"))(100))

# identify genes with the desired adjusted p-value cut-off
ONT_bottom5000.DGEgenes <- rownames(subset(ONT_bottom5000.results, padj<0.05)) 
# extract rlog-transformed values into a matrix
ONT_bottom5000_rlog.dge <- ONT_bottom5000.rlog[ONT_bottom5000.DGEgenes,] %>% assay
# heatmap of DEG sorted by p.adjust
pheatmap(ONT_bottom5000_rlog.dge, scale="row",
         show_rownames=FALSE, main="ONT DGE",
         color = colorRampPalette(c("blue", "red"))(100))

```


```{r}
median_8000_genes <- gencode.vM33.All.GeneLengths.filtered[order(gencode.vM33.All.GeneLengths.filtered$ab_diff)[1:8000],]

top_8000_genes <- gencode.vM33.All.GeneLengths.filtered[order(-gencode.vM33.All.GeneLengths.filtered$GeneLength)[1:8000],]

bottom_8000_genes <- gencode.vM33.All.GeneLengths.filtered[order(gencode.vM33.All.GeneLengths.filtered$GeneLength)[1:8000],]

sample_genes <- list(
  median500 = median_500_genes,
  top500 = top_500_genes,
  bottom500 = bottom_500_genes,
  median1000 = median_1000_genes,
  top1000 = top_1000_genes,
  bottom1000 = bottom_1000_genes,
  median5000 = median_5000_genes,
  top5000 = top_5000_genes,
  bottom5000 = bottom_5000_genes,
  median8000 = median_8000_genes,
  top8000 = top_8000_genes,
  bottom8000 = bottom_8000_genes
)

ONT_median8000.ds <- ONT_DESeq_overlap.ds[row.names(median_8000_genes), ]
ONT_top8000.ds <- ONT_DESeq_overlap.ds[row.names(top_8000_genes), ]
ONT_bottom8000.ds <- ONT_DESeq_overlap.ds[row.names(bottom_8000_genes), ]

ONT_median8000.rlog <- rlog(ONT_median8000.ds,blind=TRUE)
plotPCA(ONT_median8000.rlog, intgroup="condition", ntop = 100, returnData=FALSE)

ONT_top8000.rlog <- rlog(ONT_top8000.ds,blind=TRUE)
plotPCA(ONT_top8000.rlog, intgroup="condition", ntop = 100, returnData=FALSE)

ONT_bottom8000.rlog <- rlog(ONT_bottom8000.ds,blind=TRUE)
plotPCA(ONT_bottom8000.rlog, intgroup="condition", ntop = 100, returnData=FALSE)

design(ONT_median8000.ds) <- ~ condition
ONT_median8000.ds$condition %<>% relevel(ref="GC")
ONT_median8000.ds %<>% DESeq()
# normalize for diffs in sequencing depth and abundance per sample
ONT_median8000.ds %<>% estimateSizeFactors()
# gene-wise dispersion estimates across all samples
ONT_median8000.ds %<>% estimateDispersions()
# fit a neg. binomial GLM and compute Wald stat for each gene 
ONT_median8000.ds %<>% nbinomWaldTest()

ONT_median8000.results <- results(ONT_median8000.ds,independentFiltering = TRUE,alpha = 0.05)

design(ONT_top8000.ds) <- ~ condition
ONT_top8000.ds$condition %<>% relevel(ref="GC")
ONT_top8000.ds %<>% DESeq()
# normalize for diffs in sequencing depth and abundance per sample
ONT_top8000.ds %<>% estimateSizeFactors()
# gene-wise dispersion estimates across all samples
ONT_top8000.ds %<>% estimateDispersions()
# fit a neg. binomial GLM and compute Wald stat for each gene 
ONT_top8000.ds %<>% nbinomWaldTest()

ONT_top8000.results <- results(ONT_top8000.ds,independentFiltering = TRUE,alpha = 0.05)


design(ONT_bottom8000.ds) <- ~ condition
ONT_bottom8000.ds$condition %<>% relevel(ref="GC")
ONT_bottom8000.ds %<>% DESeq()
# normalize for diffs in sequencing depth and abundance per sample
ONT_bottom8000.ds %<>% estimateSizeFactors()
# gene-wise dispersion estimates across all samples
ONT_bottom8000.ds %<>% estimateDispersions()
# fit a neg. binomial GLM and compute Wald stat for each gene 
ONT_bottom8000.ds %<>% nbinomWaldTest()

ONT_bottom8000.results <- results(ONT_bottom8000.ds,independentFiltering = TRUE,alpha = 0.05)

# identify genes with the desired adjusted p-value cut-off
ONT_median8000.DGEgenes <- rownames(subset(ONT_median8000.results, padj<0.05)) 
# extract rlog-transformed values into a matrix
ONT_median8000_rlog.dge <- ONT_median8000.rlog[ONT_median8000.DGEgenes,] %>% assay
# heatmap of DEG sorted by p.adjust
pheatmap(ONT_median8000_rlog.dge, scale="row",
         show_rownames=FALSE, main="ONT DGE",
         color = colorRampPalette(c("blue", "red"))(100))

# identify genes with the desired adjusted p-value cut-off
ONT_top8000.DGEgenes <- rownames(subset(ONT_top8000.results, padj<0.05)) 
# extract rlog-transformed values into a matrix
ONT_top8000_rlog.dge <- ONT_top8000.rlog[ONT_top8000.DGEgenes,] %>% assay
# heatmap of DEG sorted by p.adjust
pheatmap(ONT_top8000_rlog.dge, scale="row",
         show_rownames=FALSE, main="ONT DGE",
         color = colorRampPalette(c("blue", "red"))(100))

# identify genes with the desired adjusted p-value cut-off
ONT_bottom8000.DGEgenes <- rownames(subset(ONT_bottom8000.results, padj<0.05)) 
# extract rlog-transformed values into a matrix
ONT_bottom8000_rlog.dge <- ONT_bottom8000.rlog[ONT_bottom8000.DGEgenes,] %>% assay
# heatmap of DEG sorted by p.adjust
pheatmap(ONT_bottom8000_rlog.dge, scale="row",
         show_rownames=FALSE, main="ONT DGE",
         color = colorRampPalette(c("blue", "red"))(100))
```


### Length Composition Analysis

```{r}
gencode.vM33.ONT.GeneLengths <- gencode.vM33.All.GeneLengths[row.names(ONT_DESeq.ds),]
gencode.vM33.ONT.GeneLengths.overlap <- gencode.vM33.All.GeneLengths.filtered[row.names(ONT_DESeq_overlap.ds),]

hist(gencode.vM33.ONT.GeneLengths$log2.GeneLength, freq = FALSE)

ONT_medians <- apply(counts(ONT_DESeq.ds), 1, median)

gencode.vM33.ONT.GeneLengths$medianCounts <- ONT_medians[match(rownames(gencode.vM33.ONT.GeneLengths), names(ONT_medians))]
gencode.vM33.ONT.GeneLengths.overlap$medianCounts <- ONT_medians[match(rownames(gencode.vM33.ONT.GeneLengths.overlap), names(ONT_medians))]

#Scatter plot of log(medianCounts) and log(geneLengths)
plot(log2(medianCounts) ~ log2.GeneLength, data = gencode.vM33.ONT.GeneLengths, 
     main = "Scatter Plot of log2(Median Counts) vs. log2(Gene Length)",
     xlab = "log2(Gene Length)", ylab = "log2(Median Counts)", col = "black", pch = 16)

```

```{r}
gencode.vM33.Illumina.GeneLengths <- gencode.vM33.All.GeneLengths[row.names(Illumina_DESeq.ds),]
gencode.vM33.Illumina.GeneLengths.overlap <- gencode.vM33.All.GeneLengths.filtered[row.names(Illumina_DESeq_overlap.ds),]

hist(gencode.vM33.Illumina.GeneLengths$log2.GeneLength, freq = FALSE)

Illumina_medians <- apply(counts(Illumina_DESeq.ds), 1, median)

gencode.vM33.Illumina.GeneLengths$medianCounts <- Illumina_medians[match(rownames(gencode.vM33.Illumina.GeneLengths), names(Illumina_medians))]
gencode.vM33.Illumina.GeneLengths.overlap$medianCounts <- Illumina_medians[match(rownames(gencode.vM33.Illumina.GeneLengths.overlap), names(Illumina_medians))]

#Scatter plot of log(medianCounts) and log(geneLengths)
plot(log2(medianCounts) ~ log2.GeneLength, data = gencode.vM33.Illumina.GeneLengths, 
     main = "Scatter Plot of log2(Median Counts) vs. log2(Gene Length)",
     xlab = "log2(Gene Length)", ylab = "log2(Median Counts)", col = "black", pch = 16)
```

```{r}
ONT_hist1 <- hist(gencode.vM33.ONT.GeneLengths$log2.GeneLength, freq = FALSE)
Illumina_hist1 <- hist(gencode.vM33.Illumina.GeneLengths$log2.GeneLength, freq = FALSE)

# Set the alpha (transparency) value, e.g., 0.5 for 50% transparency
transparency <- 0.5

# Create translucent colors using the rgb function
red_translucent <- rgb(1, 0, 0, alpha = transparency)
blue_translucent <- rgb(0, 0, 1, alpha = transparency)

# Find the overall range of y values
y_max <- max(c(ONT_hist1$counts, Illumina_hist1$counts))

# Plot histograms with translucent colors and adjusted y-axis limits
plot(ONT_hist1, col = red_translucent, ylim = c(0, y_max))  # first histogram
plot(Illumina_hist1, col = blue_translucent, add = TRUE)     # second histogram
```

```{r}
t.test(gencode.vM33.ONT.GeneLengths$log2.GeneLength, gencode.vM33.Illumina.GeneLengths$log2.GeneLength, alternative = "two.sided", var.equal = FALSE)
```

```{r}
#increase transparency
transparency <- 0.3
# Create translucent colors using the rgb function
red_translucent <- rgb(1, 0, 0, alpha = transparency)
blue_translucent <- rgb(0, 0, 1, alpha = transparency)

# Scatter plot of log(medianCounts) and log(geneLengths) for ONT data
plot(log2(medianCounts) ~ log2.GeneLength, data = gencode.vM33.ONT.GeneLengths.overlap,
     main = "Scatter Plot of log2(Median Counts) vs. log2(Gene Length)",
     xlab = "log2(Gene Length)", ylab = "log2(Median Counts)", col = red_translucent, pch = 1,cex=0.1)

# Add points for Illumina data on the same plot
points(log2(medianCounts) ~ log2.GeneLength, data = gencode.vM33.Illumina.GeneLengths.overlap,
       col = blue_translucent, pch = 2,cex=0.1)

```
Illumina overestimates/gives more read counts to longer genes. 

###Fit distributions of the scatter plots. 

```{r}
#Combine the median gene counts from both platforms
gencode.vM33.Illumina.GeneLengths$log2.medianCounts <- log2(gencode.vM33.Illumina.GeneLengths$medianCounts + 1)
gencode.vM33.ONT.GeneLengths$log2.medianCounts <- log2(gencode.vM33.ONT.GeneLengths$medianCounts + 1)

gencode.vM33.Illumina.GeneLengths.overlap$log2.medianCounts <- log2(gencode.vM33.Illumina.GeneLengths.overlap$medianCounts + 1)
gencode.vM33.ONT.GeneLengths.overlap$log2.medianCounts <- log2(gencode.vM33.ONT.GeneLengths.overlap$medianCounts + 1)

gencode.vM33.Combined.GeneLengths <- cbind(gencode.vM33.Illumina.GeneLengths.overlap,gencode.vM33.ONT.GeneLengths.overlap[,5:6])

selected_columns1 <- 5:6
prefix1 <- "Illumina_" 

colnames(gencode.vM33.Combined.GeneLengths)[selected_columns1] <- paste0(prefix1,colnames(gencode.vM33.Illumina.GeneLengths.overlap)[selected_columns1])

selected_columns2 <- 7:8
prefix2 <- "ONT_"
colnames(gencode.vM33.Combined.GeneLengths)[selected_columns2] <- paste0(prefix2,colnames(gencode.vM33.ONT.GeneLengths.overlap)[selected_columns1])
``` 


```{r}
hist(gencode.vM33.Combined.GeneLengths$Illumina_log2.medianCounts,freq = FALSE)
lines(density(gencode.vM33.Combined.GeneLengths$Illumina_log2.medianCounts))
hist(gencode.vM33.Combined.GeneLengths$ONT_log2.medianCounts, freq = FALSE)
lines(density(gencode.vM33.Combined.GeneLengths$ONT_log2.medianCounts))

par(mar = c(5, 5, 2, 2))
descdist(gencode.vM33.Combined.GeneLengths$Illumina_log2.medianCounts, discrete = FALSE)
descdist(gencode.vM33.Combined.GeneLengths$ONT_log2.medianCounts, discrete = FALSE)
```
```{r warning=TRUE}

cor(gencode.vM33.Combined.GeneLengths$Illumina_log2.medianCounts, gencode.vM33.Combined.GeneLengths$log2.GeneLength, method='pearson')

cor(gencode.vM33.Combined.GeneLengths$ONT_log2.medianCounts, gencode.vM33.Combined.GeneLengths$log2.GeneLength, method='pearson')
```

Since the correlations are not linear for ONT, I will try to assess the correlation assuming the data is non-linearly correlated. 

```{r}
nlcor(gencode.vM33.Combined.GeneLengths$log2.GeneLength,gencode.vM33.Combined.GeneLengths$Illumina_log2.medianCounts,plt=T)

nlcor(gencode.vM33.Combined.GeneLengths$log2.GeneLength,gencode.vM33.Combined.GeneLengths$ONT_log2.medianCounts,plt=T)
```

```{r}
transparency1 <- 0.2
transparency2 <- 0.1
# Scatter plot of log(medianCounts) and log(geneLengths) for ONT data
ggplot(gencode.vM33.ONT.GeneLengths.overlap, aes(x = log2.GeneLength, y = log2(medianCounts+1))) +
  geom_point(aes(color = "ONT"), alpha = transparency1, size = 0.5) +
  
  # Add points for Illumina data on the same plot
  geom_point(data = gencode.vM33.Illumina.GeneLengths.overlap, aes(x = log2.GeneLength, y = log2(medianCounts+1), color = "Illumina"), alpha = transparency2, size = 0.5) +
  
  # Add correlation line for Illumina data
  geom_smooth(data = gencode.vM33.Illumina.GeneLengths.overlap, method = "lm", formula = y ~ x, se = FALSE, linetype = "solid", color = "blue") +
  
  # Add correlation line for ONT data
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linetype = "solid", color = "red") +
  
  # Customize the plot
  labs(title = "Scatter Plot of log2(Median Counts) vs. log2(Gene Length)",
       x = "log2(Gene Length)", y = "log2(Median Counts)",color="platform") +
  scale_color_manual(values = c("ONT" = "red", "Illumina" = "blue")) +
  # Increase the size of the legend
  guides(color = guide_legend(override.aes = list(size = 4))) +
  theme_minimal()
```


```{r}
transparency1 <- 0.2
transparency2 <- 0.1
# Scatter plot of log(medianCounts) and log(geneLengths) for ONT data
ggplot(gencode.vM33.ONT.GeneLengths.overlap, aes(x = log2.GeneLength, y = log2(medianCounts+1))) +
  geom_point(aes(color = "ONT"), alpha = transparency1, size = 0.5) +
  
  # Add correlation line for ONT data
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linetype = "solid", color = "red") +
  
  # Customize the plot
  labs(title = "Scatter Plot of log2(Median Counts) vs. log2(Gene Length)",
       x = "log2(Gene Length)", y = "log2(Median Counts)",color="platform") +
  # Increase the size of the legend
  guides(color = guide_legend(override.aes = list(size = 4))) +
  theme_minimal()
```

```{r}
t.test()
```


There is negligible correlation between nanopore sequencing gene counts and gene length. 